---
title: "Shiny Application Project"
subtitle: "Developing Data Products - Coursera"
author: "Mayank Singh"
date: "1/19/2018"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Overview

The shiny application displays an analysis of how Horsepower is correlated to the MPG of a vehicle.

- Model 1 models the linear relationship between HP & MPG.

- Model 2 models the polynomial (cubic) relationship between HP & MPG.



## Analysis Plots

A scatter plot of Horsepower v/s Miles per Gallon
```{r cars, echo = FALSE}
data("mtcars")
model1 <- lm(hp ~ mpg, data = mtcars)
model2 <- lm(hp ~ poly(mpg, 3), data = mtcars)

plot(mtcars$mpg, mtcars$hp, xlab = "Miles Per Gallon",
     ylab = "Horsepower", bty = "n", pch = 16,
     xlim = c(10, 35), ylim = c(50, 350))


```

## Plotting Regression Models

```{r pressure}

plot(mtcars$mpg, mtcars$hp, xlab = "Miles Per Gallon",
     ylab = "Horsepower", bty = "n", pch = 16,
     xlim = c(10, 35), ylim = c(50, 350))

abline(model1, col = "red", lwd = 2)
predicted.intervals <- predict(model2, mtcars)
lines(mtcars$mpg, predicted.intervals, col='green',lwd=2)

```


## Applicability of Analysis

The analysis above can be used for variety of purposes:

- It's applicability in finding the perfect tradeoff between power and efficiency is immense.
- The above application can also be used as a supplement to a buying guide where individuals can assess the fuel economy v/s the torque needs they may have for their work.
- It can also serve as a guide for Automobile Engineering students looking to explore the relationship between HP and MPG.
