Codebook
========
Codebook was generated on `r as.character(Sys.time())` during the same process that generated the dataset. See `run_analysis.md` or `run_analysis.html` for details on dataset creation.

Variable list and descriptions
------------------------------

Variable name    | Description
-----------------|------------
subject          | ID the subject who performed the activity for each window sample. Its range is from 1 to 30.
activity         | Activity name

domain_ftr       | Feature: Time domain signal or frequency domain signal (Time or Freq)
instrument_ftr   | Feature: Measuring instrument (Accelerometer or Gyroscope)
acceleration_ftr | Feature: Acceleration signal (Body or Gravity)
variable_ftr     | Feature: Variable (Mean or SD)
jerk_ftr         | Feature: Jerk signal
magnitude_ftr    | Feature: Magnitude of the signals calculated using the Euclidean norm
axis_ftr         | Feature: 3-axial signals in the X, Y and Z directions (X, Y, or Z)
count_ftr        | Feature: Count of data points used to compute `average`
average_ftr      | Feature: Average of each variable for each activity and each subject


Dataset structure
-----------------

```{r}
str(tidy_data)
```

List the key variables in the data table
----------------------------------------

```{r}
key(tidy_data)
```

Show a few rows of the dataset
------------------------------

```{r}
tidy_data
```

Summary of variables
--------------------

```{r}
summary(tidy_data)
```

List all possible combinations of features
------------------------------------------

```{r}
tidy_data[, .N, by=c(names(tidy_data)[grep("^feat", names(tidy_data))])]
```